
---
title: 'Demo of DIRECT-NET using parallel scATAC-seq data of human GM12878 cell line'
author: "Lihua Zhang"
date: "`r format(Sys.time(), '%d/%m/%y')`"
output:
  html_document: default
mainfont: Arial
vignette: |
  %\VignetteIndexEntry{Inferring CREs and reconstructing regulatory networks from single cell multi-omics data} %\VignetteEngine{knitr::rmarkdown} %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  root.dir = './'
)
```

We showcase DIRECT-NET’s capability of inferring CREs by applying it to scATAC-seq data of human GM12878 cell line. 

To make it easy to run DIRECT-NET in most common parallel scRNA-seq and scATAC-seq data or scATAC-seq data analysis pipelines, the implementent of DIRECT-NET is seamlessly compatible with the workflow of Seurat/Signac package. In sum, DIRECT-NET pipeline consists of the following four major parts:

* Set up a Seurat object

* Infer links using DIRECT-NET by taking an input a Seurat object

* Visualize the links

* Construct regulatory networks

DIRECT-NET's workflow closely follows the Seurat vignette: [https://satijalab.org/seurat/articles/weighted_nearest_neighbor_analysis.html](https://satijalab.org/seurat/articles/weighted_nearest_neighbor_analysis.html)

## Load the required libraries
```{r message=FALSE,warning=FALSE}
library(DIRECTNET)
library(Signac)
library(Seurat)
library(S4Vectors)
library(patchwork)
library(Matrix)
library(data.table)
options(stringsAsFactors = FALSE)
set.seed(123)
```

## Load data
The scATAC-seq datasets we demonstrated here, including count data matrix of GM12878, can be downloaded via this shared Google Drive [link](https://drive.google.com/drive/folders/1--kXLO7FS5mRN5-aVjPV6WmS4Dcos4b3?usp=sharing).

```{r}
load("/Users/lihuazhang/Documents/DIRECT-NET/GM12878/Data/GM12878.filtered.cells.used.for.linkage.analysis.RData")
atac.counts <- t(as.matrix(mat_ct_filtered))
features.rna <- read.table(file = "/Users/lihuazhang/Documents/DIRECT-NET/GM12878/Data/GM12878_hvg.txt")
features.rna <- features.rna$x
focus_markers <- features.rna

genome.info <- read.table(file = "/Users/lihuazhang/Documents/DIRECT-NET/GM12878/Data/hg19.promoter.regions.txt")
names(genome.info) <- c("Chrom","Starts","Ends","genes")
genes <- lapply(genome.info$genes, function(x) strsplit(x,"[|]")[[1]][1])
genes <- lapply(genes, function(x) strsplit(x,"[.]")[[1]][1])
genes <- unlist(genes)
genome.info$genes <- genes
unik <- !duplicated(genes)# filter out different transcript
genome.info <- genome.info[unik,]

peak <- rownames(atac.counts)
peak1 <- strsplit(peak,"_")
peaks <- matrix(0, nrow = length(peak),ncol = 1)
for (i in 1:length(peaks)) {
  peaks[i] <- paste0(peak1[[i]][1],":",peak1[[i]][2],"-",peak1[[i]][3])
}
rownames(atac.counts) <- as.vector(peaks)

```

# Part I: Set up a Seurat object
```{r, warning=FALSE}
sci <- CreateSeuratObject(
  counts = atac.counts,
  assay = "ATAC"
)

```

### ATAC analysis
```{r}
# We exclude the first dimension as this is typically correlated with sequencing depth
sci <- RunTFIDF(sci)
sci <- FindTopFeatures(sci, min.cutoff = 50)
sci <- RunSVD(sci)
DepthCor(sci)
sci <- RunUMAP(sci, reduction = 'lsi', dims = 2:30)
sci
sci <- FindNeighbors(object = sci, reduction = 'lsi', dims = 2:30)
sci <- FindClusters(object = sci, verbose = FALSE, algorithm = 3, resolution = 0.2)
#DimPlot(object = sci, label = TRUE) + NoLegend()

```

# Part II: Infer CREs using DIRECT-NET
```{r,warning=FALSE}
sci <- Run_DIRECT_NET(sci, peakcalling = FALSE, k_neigh = 50, atacbinary = TRUE, max_overlap=0.5, size_factor_normalize = TRUE, genome.info = genome.info, focus_markers = focus_markers[1:5])
direct.net_result <- Misc(sci, slot = 'direct.net')
direct.net_result <- as.data.frame(do.call(cbind,direct.net_result))

```

# Part III: Visualize the links
```{r,warning=FALSE}
temp <- tempfile()
download.file("ftp://ftp.ensembl.org/pub/release-65/gtf/homo_sapiens/Homo_sapiens.GRCh37.65.gtf.gz", temp)
gene_anno <- rtracklayer::readGFF(temp)
unlink(temp)
# rename some columns to match requirements
gene_anno$chromosome <- paste0("chr", gene_anno$seqid)
gene_anno$gene <- gene_anno$gene_id
gene_anno$transcript <- gene_anno$transcript_id
gene_anno$symbol <- gene_anno$gene_name
marker <- focus_markers[1]
Plot_connections(direct.net_result, gene_anno, marker)
```
