
---
title: 'Demo of DIRECT-NET using scATAC-seq data of human Brain'
author: "Lihua Zhang"
date: "`r format(Sys.time(), '%d/%m/%y')`"
output:
  html_document: default
mainfont: Arial
vignette: |
  %\VignetteIndexEntry{Inferring CREs and reconstructing regulatory networks from single cell multi-omics data} %\VignetteEngine{knitr::rmarkdown} %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  root.dir = './'
)
```

We showcase DIRECT-NET’s capability of inferring CREs by applying it to scATAC-seq data of human Brain. 

To make it easy to run DIRECT-NET in most common parallel scRNA-seq and scATAC-seq data or scATAC-seq data analysis pipelines, the implementent of DIRECT-NET is seamlessly compatible with the workflow of Seurat/Signac package. In sum, DIRECT-NET pipeline consists of the following four major parts:

* Set up a Seurat object

* Infer links using DIRECT-NET by taking an input a Seurat object

* Visualize the links

* Construct regulatory networks

DIRECT-NET's workflow closely follows the Seurat vignette: [https://satijalab.org/seurat/articles/weighted_nearest_neighbor_analysis.html](https://satijalab.org/seurat/articles/weighted_nearest_neighbor_analysis.html)

## Load the required libraries
```{r message=FALSE,warning=FALSE}
library(DIRECTNET)
library(Seurat)
library(Signac)
library(EnsDb.Hsapiens.v86)
library(patchwork)
library(dplyr)
library(ggplot2)
options(stringsAsFactors = FALSE)
```

## Load data
The scATAC-seq datasets we demonstrated here can be downloaded via this shared Google Drive [link](https://drive.google.com/drive/folders/1ff3HD7x_Rv5v1y1zhbZDqjMeXMJDMnaU?usp=sharing).

```{r}
library(SummarizedExperiment)
Data <- readRDS(file = "/Users/lihuazhang/Documents/DIRECT-NET/Brain/Data/GSE147672_200324_Brain_scATAC_SummarizedExperiment_forGEO")
counts <- assay(Data, "counts")
counts <- counts[which(rowSums(counts) > 1000),]
meta = as.data.frame(colData(Data))
# import low dimension data and cluster information
library(readxl)
Umap_info <- read_excel("/Users/lihuazhang/Documents/DIRECT-NET/Brain/Data/41588_2020_721_MOESM5_ESM.xlsx",skip = 21)
#head(Umap_info)
cell_coord <- cbind(Umap_info$UMAP_dim1,Umap_info$UMAP_dim2)
#head(cell_coord)
rownames(cell_coord) <- paste0("scATAC_CTRL_",Umap_info$Region,"_",Umap_info$Donor_ID,"#",Umap_info$`10x_SingleCell_Barcode`,"-1")
colnames(cell_coord) <- c("UMAP_1","UMAP_2")
index <- match(colnames(counts),rownames(cell_coord))
identical(colnames(counts),rownames(cell_coord)[index])
cell_coord <- cell_coord[index,]
identical(colnames(counts),rownames(cell_coord))
group0 <- Umap_info$Cluster[index]
#unique(group0)
# based on original nature genetics paper (https://www.nature.com/articles/s41588-020-00721-x)
group <- c('Striatal inhibitory 1', "Oligodendrocytes 2",'Striatal inhibitory 2',"Oligodendrocytes 1","unclassified","Oligodendrocytes 4","Isocortical exitatory","OPCs 2","unclassified","Oligodendrocytes 5","Microglia","unclassified","Oligodendrocytes 3","OPCs 1","unclassified","Hippocampal exitatory 2","Isocortical astrocytes","Striatal astrocytes","Hippocampal exitatory 1","Isocortical astrocytes","unclassified",'Isocortical inhibitory',"unclassified","Nigral OPCs")
group_new <- plyr::mapvalues(group0, from = unique(group0), to = group)
#group_new
index <- which(group_new != "unclassified")
counts <- counts[,index]
meta <- meta[index,]
meta$celltype <- group_new[index]

genome.info <- read.table(file = "/Users/lihuazhang/Documents/DIRECT-NET/Brain/Data/hg38.promoter.regions.txt")
names(genome.info) <- c("Chrom","Starts","Ends","genes")
genes <- lapply(genome.info$genes, function(x) strsplit(x,"[|]")[[1]][1])
genes <- lapply(genes, function(x) strsplit(x,"[.]")[[1]][1])
genes <- unlist(genes)
genome.info$genes <- genes
unik <- !duplicated(genes)# filter out different transcript
genome.info <- genome.info[unik,]

markers <- read.table("/Users/lihuazhang/Documents/DIRECT-NET/Brain/Data/Brain_markers.txt", sep = "\t")
markers <- data.frame(gene = markers$gene, group = markers$cluster)
c <- unique(markers$group)
marker_list <- list()
for (i in 1:length(c)) {
  marker1<- markers[markers$group == c[i],]
  marker_list[[i]] <- as.character(marker1$gene)
}

markers_all <- unique(unlist(marker_list))
focus_markers <- markers_all
```

# Part I: Set up a Seurat object
```{r, warning=FALSE}
combined <- CreateSeuratObject(
  counts = counts,
  meta.data = meta,
  assay = "ATAC"
)

```

### ATAC analysis
```{r}
# We exclude the first dimension as this is typically correlated with sequencing depth
combined <- RunTFIDF(combined)
combined <- FindTopFeatures(combined, min.cutoff = 50)
combined <- RunSVD(combined)
DepthCor(combined)
combined <- RunUMAP(combined, reduction = 'lsi', dims = 2:30)
Idents(combined) <- combined$celltype
combined@reductions$umap@cell.embeddings <- cell_coord

```

# Part II: Infer CREs using DIRECT-NET
```{r,message=FALSE,warning=FALSE}
combined <- Run_DIRECT_NET(combined, peakcalling = FALSE, k_neigh = 50, atacbinary = TRUE, max_overlap=0.5, size_factor_normalize = TRUE, genome.info = genome.info, focus_markers = "FEV")
direct.net_result <- Misc(combined, slot = 'direct.net')
direct.net_result <- as.data.frame(do.call(cbind,direct.net_result))
# We have run DIRECT-NET on all markers, load the results for downstream network analysis
load("/Users/lihuazhang/Documents/DIRECT-NET/Brain/Data/Brain_direct.net.RData")
# check the function type name
direct.net_result$function_type <- gsub("HF","HC",direct.net_result$function_type)
direct.net_result$function_type <- gsub("Rest","MC",direct.net_result$function_type)
direct.net_result$function_type <- gsub("LF","LC",direct.net_result$function_type)
```

# Part III: Visualize the links
```{r,warning=FALSE}
temp <- tempfile()
download.file("ftp://ftp.ensembl.org/pub/release-100/gtf/homo_sapiens/Homo_sapiens.GRCh38.100.gtf.gz", temp)
gene_anno <- rtracklayer::readGFF(temp)
unlink(temp)
# rename some columns to match requirements
gene_anno$chromosome <- paste0("chr", gene_anno$seqid)
gene_anno$gene <- gene_anno$gene_id
gene_anno$transcript <- gene_anno$transcript_id
gene_anno$symbol <- gene_anno$gene_name
Plot_connections(direct.net_result, gene_anno, marker = "FEV", cutoff = 0.2)
```

# Part IV: Construct regulatory networks
# identify differential accessible peaks, here we focus on Striatal inhibitory 1 and Striatal inhibitory 2
```{r, message=FALSE,warning=FALSE}
### identify differential accessible peaks (DA)
DefaultAssay(combined) <- 'ATAC'
focused_markers <- markers[which(markers$group %in% c("Striatal inhibitory 1", "Striatal inhibitory 2")), , drop = FALSE]
groups <- unique(focused_markers$group)
da_peaks_list <- list()
for (i in 1:length(groups)) {
  print(i)
  da_peaks <- FindMarkers(
    object = combined,
    min.pct = 0.2,
    ident.1 = groups[i],
    group.by = "celltype",
    test.use = 'LR',
    only.pos = TRUE
  )
  da_peaks_list[[i]] <- da_peaks
}
```

# Detect CRE-TF connections
```{r, message=FALSE,warning=FALSE}
# CRE-gene connections
CREs_Gene <- generate_CRE_Gene_links(direct.net_result, markers = focused_markers)
# Find focused CREs which is overlapped with DA
Focused_CREs <- generate_CRE(L_G_record = CREs_Gene$distal, P_L_G_record = CREs_Gene$promoter, da_peaks_list)
# detect TFs for distal CREs
library(BSgenome.Hsapiens.UCSC.hg38)
L_TF_record <- generate_peak_TF_links(peaks_bed_list = Focused_CREs$distal, species="Homo sapiens", genome = BSgenome.Hsapiens.UCSC.hg38, markers = focused_markers)
# detect TFs for Promoters
P_L_TF_record <- generate_peak_TF_links(peaks_bed_list = Focused_CREs$promoter, species="Homo sapiens", genome = BSgenome.Hsapiens.UCSC.hg38, markers = focused_markers)

```

# Output TF-gene connections
```{r, message=FALSE,warning=FALSE}
network_links <- generate_links_for_Cytoscape(L_G_record = CREs_Gene$distal, L_TF_record, P_L_G_record = CREs_Gene$promoter, P_L_TF_record,groups)
```

# Ouput Node attribute
```{r, message=FALSE,warning=FALSE}
Node_attribute <- generate_node_for_Cytoscape(network_links,markers = focused_markers)
```
